Statistical framework for whole-genome bisulfite-sequencing data

### 2014-12-15 Rmarkdown/datasets.Rmd
Misc. editing

This chapter briefly documents the 51 whole-genome bisulfite-sequencing samples that I use in my thesis. The `BAM` files for each sample underwent the same basic processing:

@@

__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e., going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $275$ GB alone.

@@

- _BUF_: Buffy coat layer, which are leukocytes and platelets derived by centrifugation of a whole blood sample.
- _SA_: Subcutaneous adidose tissue, which is fat found just below the skin. Unlike visceral fat, "subcutaneous fat is not related to many of the classic obesity-related pathologies, such as heart disease, cancer, and stroke" (__TODO: Get better reference than wikipedia__).
- _VA_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs (stomach, liver, intestines, kidneys, etc.) (__TODO: Get better reference than wikipedia__).
- _VAT_: Visceral adipocytes, which are derived from the VA tissue.

@@

The Ziller are a subset of the data used in \cite{Ziller:2013fs}. Specifically, I use a convenience sample of 19 (__TODO: Check final numbers__) WGBS libraries. These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research), who already had aligned and processed BAM files for these 19 libraries. These libraries in fact represent 19 sequencing runs of eight biological samples, which I was not aware of until the data had already been processed. Therefore, I have _post hoc_ merged the processed data from sequencing runs. This unfortunately reduces the power to detect genetic variants since the coverage of individual sequencing runs is lower than merged data, but does not adversely affect methylation calling since the number of methylated and unmethylated reads can be summed across sequencing runs.

>>>>

This chapter briefly documents the 40 whole-genome bisulfite-sequencing samples that I use in my thesis. The `BAM` files for each sample underwent the same basic processing:

@@

__TODO__: I'd like to make available all the data and scripts necessary to reproduce my analyses, i.e., going from a `BAM` file to a `MethPat` objects, but it's a little daunting --- the folder containing __just the processed data__ (`/home/users/lab0605/hickey/methylation_m-tuples`) is $> 275$ GB alone.

@@

- _BUF_: Buffy coat layer, which are leukocytes and platelets derived by centrifugation of a whole blood sample.
- _SA_: Subcutaneous adidose tissue, which is fat found just below the skin. Unlike visceral adipose tissue, subcutaneous adipose tissue is thought to be protective against obesity-related metabolic dysfunction \citep{Chau:2014do}.
- _VA_: Visceral adipose tissue, which is located inside the abdominal cavity, packed between the organs and is associated with metabolic dysfunction \citep{Chau:2014do}.
- _VAT_: Visceral adipocytes, which are derived from the VA tissue.

@@

The Ziller are a subset of the data used in \cite{Ziller:2013fs}. Specifically, I use a convenience sample of 8 WGBS libraries. These were made available to me by a collaborator, Aaron Statham (Garvan Institute of Medical Research), who already had aligned and processed BAM files for these 19 libraries. These libraries in fact represent 19 sequencing runs of eight biological samples, which I was not aware of until the data had already been processed. Therefore, I have _post hoc_ merged the processed data from sequencing runs. This unfortunately reduces the power to detect genetic variants since the coverage of individual sequencing runs is lower than merged data, but does not adversely affect methylation calling since the number of methylated and unmethylated reads can be summed across sequencing runs.

### 2014-12-15 Rmarkdown/wgbs_statistical_framework.Rmd
Editing
Misc. editing

\subsubsection{Genetic heterozygosity at a methylation locus}\label{sec:heterozygosity}

@@

__TODO: Update discussion of results in light of Ziller merged data__

@@

Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively. Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_merged_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.

What is immediately clear is that these distributions are bimodal. The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.

Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _SRR949208_ and _SRR949209_ from the Ziller dataset --- or cancer cells lines and tissue --- _SRR949206_, _SRR949207_, _SRR949210_, _SRR949211_, _SRR949212_, _SRR949213_ and _SRR949215_ from the Ziller dataset. Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the Seisenberger dataset also displays greater levels of intermediate methylation. This sample was also created by pooling DNA from multiple mice, which may explain the extra variability in the $\beta$-value distribution.

@@

The explanation for the somatic cell lines is less clear. Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation. Likewise, the various frontal cortex samples in the Ziller dataset --- _SRR949193_, _SRR949194_, _SRR949195_, _SRR949196_, _SRR949197_, _SRR949198_, _SRR949199_, _SRR949201_, _SRR949202_, _SRR949203_ and _SRR949205_ --- which includes both 'normal' and Alzheimer's samples, have very low levels of intermediate methylation. This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_. Naively, a cell line is a 'pure' cell population, however, the DNA methylation data clearly reveal widespread differences between cells in these populations.

@@

In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data. I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured. By using a common statistical framework we can better understand how different statistical methods relate to one another. Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties. I then examined the empirical distributions of these variables across a diverse set of 51 whole-genome bisulfite-sequencing samples.

Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA. Nonetheless, some of the increased variability in the Seisenberger data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.

>>>>

\subsubsection{Genetic heterozygosity at a methylation locus} \label{sec:heterozygosity}

@@

@@

Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity}, \ref{fig:Lister_plotBetaDensity}, \ref{fig:Seisenberger_plotBetaDensity} and \ref{fig:Ziller_merged_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberger and Ziller datasets, respectively. Figures \ref{fig:EPISCOPE_plotBetaFreqPoly}, \ref{fig:Lister_plotBetaFreqPoly}, \ref{fig:Seisenberger_plotBetaFreqPoly} and \ref{fig:Ziller_merged_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.

What is immediately clear is that these distributions are bimodal --- most CpGs are highly methylated or lowly methylated. The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.

Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _IMR90\_cell\_line_ from the Ziller dataset --- or cancer cells lines and tissue --- _HepG2\_cell\_line_, _Colon\_Tumor\_Primary_ and _Colon\_Primary\_Normal_ from the Ziller dataset. Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the Seisenberger dataset also displays greater levels of intermediate methylation. This sample was also created by pooling DNA from multiple mice, which may explain the extra variability in the $\beta$-value distribution.

@@

The explanation for the somatic cell lines is less clear. Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation. Likewise, the various frontal cortex samples in the Ziller dataset --- _Frontal\_cortex\_normal\_1_, _Frontal\_cortex\_normal\_2_, _Frontal\_cortex\_AD\_1_ and _Frontal\_cortex\_AD\_2_ --- which includes both 'normal' and Alzheimer's samples, have very low levels of intermediate methylation. This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_. Naively, a cell line is a 'pure' cell population, however, the DNA methylation data clearly reveal widespread differences between cells in these populations.

@@

In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data. I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured. By using a common statistical framework we can better understand how different statistical methods relate to one another. Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties. I then examined the empirical distributions of these variables across a diverse set of 40 whole-genome bisulfite-sequencing samples.

Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA. Nonetheless, some of the increased variability in the Seisenberger data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.
